Multisequence algorithm for coarse-grained biomolecular simulations: application to protein fold switching

### 2017-06-01 Manuscript.tex
Miscellaneous changes

\newcommand	 {\sbar}	{{s}}
\newcommand	 {\rbar}	{{r}}
\newcommand	 {\hi}		{{h_\mathrm{i}}}
\newcommand	 {\pii}  	{{p_\mathrm{i}}}
\newcommand	 {\kb}		{{k_\mathrm{B}}}
\newcommand	 {\Tlow}	{{T_\mathrm{low}}}
\newcommand 	 {\Pnat} 	{{P_\mathrm{nat}}}
\newcommand {\QIA}	{{Q_\mathrm{IA}}}
\newcommand {\QIB}	{{Q_\mathrm{IB}}}
\newcommand {\QIIA}	{{Q_\mathrm{IIA}}}
\newcommand {\QIIB}	{{Q_\mathrm{IIB}}}
\newcommand {\Pcut}     	{{P_\mathrm{cut}}}
\newcommand {\TlowI}     {{T^\mathrm{I}_\mathrm{low}}}
\newcommand {\TlowII}    {{T^\mathrm{II}_\mathrm{low}}}
\newcommand {\Ptot}	{{P_\mathrm{tot}}}
\newcommand {\PIA}    	{{P_\mathrm{IA}}}
\newcommand {\PIB}    	{{P_\mathrm{IB}}}
\newcommand {\PIIA}    	{{P_\mathrm{IIA}}}
\newcommand {\PIIB}    	{{P_\mathrm{IIB}}}
\newcommand {\Eave}	{{E_\mathrm{ave}}}
\newcommand {\sigE}	{{\sigma_{\left < E \right >}}}
\newcommand {\SR}		{${\mathrm{S16}_{144}}$}
\newcommand {\SI}		{${\mathrm{S16}_{1024}}$}	
\newcommand {\SII}		{${\mathrm{S16}_{1024}}$}

@@

\begin{abstract}

We consider a generalized-ensemble algorithm for coarse-grained simulations of biomolecules, which allows the thermodynamic behaviors of two or more sequences to be determined in a single multisequence run. By carrying out a random walk in sequence space, the method also enhances conformational sampling. Escape from local energy minima is accelerated by visiting sequences for which the minima are more shallow or absent. We apply the method alongside an intermediate-resolution coarse-grained model for protein folding with 7 atoms per amino acids. The potential for large-scale coverage of sequence space is explored by applying the method to two sets with $>$1,000 sequences each. The resulting thermodynamic data is used to explore the sequence-structure and sequence-stability relationships between pairs of protein folds with different secondary structures. The potential implications for the evolution of new protein folds are discussed.

\end{abstract}

\pacs{Add  here}

\keywords{Generalized ensembles, simulated tempering, protein folding, coarse-graining, evolution}

@@

\section{Introduction}
\noindent
Recent years have seen important advances in biomolecular simulation methods, including improvements to standard molecular dynamics force fields,~\cite{Piana2014} the advent of several alternative atomistic simulation approaches,~\cite{Ding2008,Irback2006,Verma2009,Yang2007} and new techniques for  conformational sampling.~\cite{Bernardi2015} These advances have in part triggered major efforts~\cite{McGuffee2010,Perilla2016,Lindorff-Larsen2011,Yu2016} to characterize the molecular dynamics of bio-systems of different sizes, e.g., a small native protein on the millisecond scale~\cite{Lindorff-Larsen2011} or a comprehensive model cytoplasm on the nanosecond scale.~\cite{Yu2016} While encouraging, these large-scale simulations also highlight the fact that severe tradeoffs in size and time scales will likely persist for the foreseeable future.

One way to expand the range of biomolecular simulations is to turn to coarse-grained models, with the basic aim to simplify the physical description of interactions while retaining the essential physics of the system under study.~\cite{Riniker2012} Ingolfsson \textit{et al.} list four main factors that make CG models computationally fast: reduction in the number of degrees of freedom, faster simulation dynamics, emphasis on short-range interactions and the ability of using larger integration time steps.~\cite{Ingolfsson2014} To this list can be added that CG representations often open up for alternative sampling schemes beyond traditional molecular dynamics approaches which can further speed up conformational sampling, e.g., activation-relaxation kinetics,~\cite{Beland2011} discrete molecular dynamics~\cite{Proctor2011} and various Monte Carlo-based strategies such as cluster moves.~\cite{Vitalis2009}

The challenges of simulating individual biomolecular systems notwithstanding, a range of systems naturally call for the investigation and comparison of many molecular variants, e.g., understanding the molecular mechanisms of specificity in protein-protein, peptide-protein,~\cite{Zarrinpar2003} or protein-nucleotide interactions, or the role of mutations in disease processes such as protein aggregation. Another example is protein folding where unique insight has been achieved by comparing the folding within and between protein families.~\cite{Tzul2017,Wensley2010}

In this work, we evaluate a Monte Carlo (MC)-based algorithm that can in principle calculate the thermodynamics of multiple sequences in a single run and apply it to a coarse-grained model for protein folding. This ``multisequence" (MS) method was originally developed in the context of homo- and heteropolymer simulations~\cite{Irback1995} and was later adapted for the characterization of peptide-protein binding specificity,~\cite{Bhattacherjee2013,Wallin2017}, but it has to our knowledge not been previously tested in protein folding simulations. The MS algorithm relies on sampling at fixed temperature a generalized ensemble in which the sequence is a dynamic parameter. Hence, there are two main types of updates: conformational updates $\rbar\rightarrow\rbar'$ and sequences updates $\sbar\rightarrow\sbar'$. As illustrated in Fig.~1, this approach requires that $\rbar$ and $\sbar$ are ``perpendicular" coordinates such that the potential energy of the model can be written in terms of two independent variables, $E(\sbar,\rbar)$. While this may be hard to achieve for detailed protein models, it is likely possible in many CG models.

As an initial test case, we apply the MS method to a set of 144 model sequences with 16 amino acids. This set, {\SR}, was constructed to sparsely cover the sequence space between two ideally designed sequences, A1 and N1, which fold into an $\alpha$-helix and $\beta$-hairpin, respectively, and allowed us to demonstrate the existence of A1-N1 mutational pathways which abruptly switch between the two structures and do not proceed through unstable sequences.~\cite{Holzgrafe2014} Here we use this set to validate the MS method and compare its efficiency to a standard generalized-ensemble methods, namely simulated tempering (ST).~\cite{Marinari1992,Lyubartsev1992} We thereafter apply the MS method to a greatly enlarged set with 1,024 sequences, also spanning the A1-N1 space, as well as another set spanning two 35-amino acid sequences, A2 and TN, folding into a two-helical bundle fold and a mixed $\alpha$-$\beta$ fold, respectively. Besides hinting at the feasibility of applying the MS method on large sets of sequences, the results allow us to carry out a more systematic analysis of the biophysical properties of sequences along mutational pathways connecting these two pairs of basic protein folds than has been possible with standard methods.~\cite{Holzgrafe2014,Holzgrafe2015}

@@

\subsection{Model sequences}

Six model sequences studied in this work are summarized in Table~I. In addition, we study two sets of 1,024 sequences derived from the A1-N1 and A2-TN pairs (see text) denoted $\mathrm{S16}_{1024}$  and $\mathrm{S35}_{1024}$, respectively, and a restricted set $\mathrm{S16}_{144}$ with 144 sequences taken from $\mathrm{S16}_{1024}$ ($\mathrm{S16}_{144}$ was also studied in Ref.~\protect\citenum{Holzgrafe2014}).

@@

\subsection{Monte Carlo simulation parameters}
\noindent
Both ST and MS simulations are carried out with two types of conformational updates $\rbar\rightarrow\rbar'$: (1) a global pivot move (20\%) which randomly picks a $\phi_\mathrm{i}$ angle or $\psi_\mathrm{i}$ angle and assign a new value between $-\pi$ and $\pi$; and (2) a semi-local move (80\%) which turns the $\phi_\mathrm{i}$ and $\psi_\mathrm{i}$-angles of 4 consecutive amino acids in a coordinated manner.~\cite{Favrin2001} In MS simulations, sequence updates $\sbar\rightarrow\sbar'$ are carried out by randomly picking a new sequence $\sbar'\ne\sbar$ and applying the accept-reject criterion in Eq.~\ref{accrej}. A sequence move is attempted every $1,000$ MC steps while temperature updates $m\rightarrow m'$ are attempted every 100 steps. The simulations carried out in this work are summarized in Tables~II.

\begin{table}
\caption{\label{tab2} List of simulations carried out in this work. }
\begin{ruledtabular}
\begin{tabular}{lccccr}
Runs & Algorithm & $\kb T$  & MC steps/run\footnote{Excludes thermalization with $10^6$ MC steps.} &  Sequences\\
\hline
32 & ST & 0.43--0.65 & $10^7$ &A1\\ 
32 & ST & 0.43--0.65 & $10^7$ &N1\\ 
32 & ST & 0.43--0.65 & $10^7$ &R1\\ 
32 & ST & 0.43--0.65 & $10^7$ &R2\\ 
32$\times$8\footnote{32  runs per temperature for 8 temperatures in the range given.} & MS &0.43--0.65& $18\times 10^7$ & $\mathrm{S16}_{144}$\\
16 & MS & 0.43  & $5\times 10^9$ &  $\mathrm{S16}_{1024}$ \\
16 & MS & 0.46 & $4\times 10^9$ &  $\mathrm{S35}_{1024}$ \\
\end{tabular}
\end{ruledtabular}
\end{table}

\subsection{Observables}
\noindent
Fold stabilities are calculated as in Ref.~\citenum{Holzgrafe2015} and also described briefly below. We define first two structural similarity measures $\QIA$ and $\QIB$ for folds IA and IB, respectively, indicating what fraction of the fold-specific backbone-backbone hydrogen bonds have been formed. The fold IA-hydrogen bonds are (2,6), (3,7), (4,8), (5,9), (6,10), (7,11), (8,12), (9,13), (10,14), (11,15) and the fold IB-bonds are (3,14), (5,12), (7,10), (10,7), (12,5), (14,3), where (i,j) indicates a hydrogen bond between the CO group of amino acid i and the NH group of amino acid j. The  stabilities of fold IA and IB are then defined as the probabilities $\PIA = P(\QIA\ge0.8)$ and $\PIB = P(\QIB\ge0.80)$, respectively, i.e., stability is the probability that at least 80\% of the fold's hydrogen bonds are formed. $\PIA$ and $\PIB$ depend on both sequence $\sbar$ and temperature $T$. For example, $\PIA=0.xx$ for A1 and $\PIB=0.xx$ for N1 at $\kb T = 0.43$. Structural similarity measures for 35-amino acid folds IIA and IIB are defined as $\QIIA = ( Q_\mathrm{IA}^\mathrm{1-16} + Q_\mathrm{IA}^\mathrm{20-35} + Q_\mathrm{tert}) ) / 3$ and $\QIIB = ( Q_\mathrm{IA}^\mathrm{1-16} + Q_\mathrm{IB}^\mathrm{20-35} + Q_\mathrm{tert} ) / 3$, respectively, where superscripts on $\QIA$ and $\QIB$ indicate over which amino acid positions those measures are applied to within the longer 35 amino acid sequences and $Q_\mathrm{tert}$ is a measure that counts the number of $C_\beta$-$C_\beta$ contacts between the two secondary structure elements of these folds.~\cite{Holzgrafe2015} As for IA and IB, the stabilities of folds IIA and IIB are defined as $\PIIA = P(\QIIA\ge0.8)$ and $\PIIB = P(\QIIB\ge0.80)$, respectively. 
%In calculating $Q$ values, a hydrogen bond is considered formed if (i) the NHO and COH angles are both $>90^\circ$ and (ii) the HO distance is $<2.7$~{\AA}.

\begin{figure}
\includegraphics[width=10.0cm]{MCTrajFolds}
\caption{(A) Example simulation trajectory of the MS algorithm applied to the sequence set $\mathrm{S16}_{144}$. The plot shows the MC evolution of the sequence $\sbar$ (numbered 1--144), the total potential energy $E$ and the root-mean-square deviation (RMSD) calculated against fold IA (light blue) and fold IB (dark red) structures shown in (B). The simulation is carried out at $\kb T = 0.43$ corresponding to the lowest studied temperature.  (B) Representative structures of fold IA and fold IB, selected to be the minimum-energy conformations found for the sequences A1 and N1, respectively. (C) Representative fold IIA and fold IIB structures selected to be the minimum-energy conformations of A2 and TN, respectively. }
\end{figure}

%
%\begin{equation}
%Q_\mathrm{\alpha\beta\beta} = ( Q_\alpha^\mathrm{1-12} + Q_\beta^\mathrm{XX} +Q_\beta^\mathrm{20-35} + Q_\mathrm{tert} ) / 4
%\end{equation}
%
%
%\begin{equation}
%Q^*_\mathrm{\alpha\beta} = [ \max(Q_\alpha^\mathrm{1-16},Q_\beta^\mathrm{XX}) +Q_\beta^\mathrm{20-35} + Q_\mathrm{tert}) ] / 3
%\end{equation}
%

@@

\subsection{Computational efficiency of the multisequence algorithm}
\noindent
To assess the computational efficiency and properties of the multisequence algorithm, we used as a test case the set of 144 3-letter model sequences with 16 amino acids studied in Refs.\cite{Holzgrafe2014} Two of the sequences are A1 and N1 which were designed to fold into stable $\alpha$-helix and $\beta$-hairpin structures, respectively. A1 and N1 differ at 10 positions such that 10 consecutive point mutations can transform A1 into N1, and vice versa. The binary sequence space between A1 and N1 in which any combination of these mutations have been carried out, therefore contains $2^{10}=1,024$ sequences. The 144 sequences were selected from this binary space with the constraint that the total number of h-amino acids are not too high or too unevenly distributed.~\cite{Holzgrafe2014}

\begin{figure}
\includegraphics[width=8.0cm]{Stderr}
\caption{Comparing MS and ST algorithms. The statistical errors $\sigma_{\left < E\right >}$ of the average total energy $\left < E\right >$ are estimated with the two methods for the sequences (A) A1, (B) N1, (C) R1 and (D) R2. Simulation lengths are adjusted such that the number of conformations sampled per sequence and temperature is similar between in MS and ST runs. }
\end{figure}

We apply the multisequence algorithm to this set of 144 sequences across a range of temperatures (see Table~II). Figure~3 illustrates a typical simulation carried out at the lowest temperature, $\Tlow$, which is well below the folding temperature of both A1 and N1~\cite{Holzgrafe2014,Holzgrafe2015}. From the MC evolution of the total energy $E$, sequence index $\sbar$, and similarity measures $Q_A$ and $Q_B$, it is evident that visits to various sequences drive transitions into a range of structural states. For example, visits to A1 ($\sbar=1$) and N1 ($\sbar=144$) tend to coincide with visits to high-$Q_A$ and high-$Q_B$ conformations, respectively, as is necessary to generate the correct conformational ensembles for these sequences.

Sequence updates $\sbar\rightarrow\sbar'$ are carried out by picking a new sequence $\sbar'$ randomly from the entire set of allowed sequences. Therefore, a sequence update typically changes more than one amino acid position. In other words, if the step size in sequence space is $\Delta H$, then typically $\Delta H>1$. Interestingly, despite the large $\Delta H$ and low $T$, the overall (average) acceptance rate $P_\mathrm{acc}=0.xx$ for $\sbar\rightarrow\sbar'$ updates which is not unreasonably low. Higher $P_\mathrm{acc}$ can be obtained by restricting proposed moves such that $\Delta H\le\Delta H_\mathrm{max}$, as shown in Figure~3. For example, an acceptance rate close to the oft-quoted rule-of-thumb value $0.25~\cite{Gilks1996} $ is obtained with $\Delta H_\mathrm{max}=4$, and with $\Delta H_\mathrm{max} = 1$ (only point mutations), $P_\mathrm{acc}=0.xx$. For simplicity, in this work, all simulations are carried out without restrictions on how new sequences $\sbar'$ are drawn.

\begin{figure}
\includegraphics[width=7.8cm]{Pacc}
\caption{Acceptance rates for $\sbar\rightarrow\sbar'$ updates in MS simulations of the {\SR} sequence set as a function of (A) the number of changed amino acid positions  and (B) temperature. }
\end{figure}

In order to assess computational efficiency, we compare with results from simulated tempering (ST) simulations carried out on 4 of the 144 sequences, namely A1 and N1 and the two sequences R1 and R2 chosen randomly at $H=4$ and $H=6$ from A1, respectively (see Table~I). While ST provides the thermodynamics of a given sequence across the 8 temperatures $T_1, ..., T_8$ in a single run, a single MS simulation provides the thermodynamics of all 144 sequences for one $T_\mathrm{i}$. We adjust the simulation lengths for ST and MS runs such that an identical number of sampled conformations are obtained for each $\sbar$ and $T_\mathrm{i}$ combination, ensuring also that similar computational resources are used for the two algorithms.

Figure 4 shows the statistical error, $\sigE$, estimated for the average energy, $\left <E\right >$, for the 4 different sequences using ST and MS algorithms. Since an identical number of sampled conformations were obatined for each s and $T_i$ combination, we can therefore compare the statistical errors directly. Both MS and ST algorithms give the same small error at the highest T studied which implies consistent result between them. It should be noted that the highest T studied is well above the folding temperature for all the 4 sequences. Therefore, this consistent result is not particularly surprising since at high-T the free-energy landscape is relatively smooth. Hence, conformational space requires little or no difficulty to sample.

@@

Insight from the statistical measurements suggests that the MS algorithm has a similar computational efficiency at $high_T$ and even better at lower T's. The MS algorithm is able to achieve  better efficency by sampling a wider region of both conformational and sequence space. Sampling the sequence space in addition to conformational space ensures a decorrelated ensemble of states is generated.

\begin{figure*}
\rotatebox{0}{\includegraphics[width=16.6cm]{network}}
\caption{Networks of the A1-N1 (top) and A2-TN (bottom) binary sequence space showing the possible stable pathways from one fold to another. (light blue) Fold A, (dark red) Fold B, and (black) Bistable. }
\end{figure*}

>>>>

\newcommand	 {\sbar}	{{s}}
\newcommand	 {\rbar}	{{r}}
\newcommand	 {\hi}		{{h_\mathrm{i}}}
\newcommand	 {\pii}  	{{p_\mathrm{i}}}
\newcommand	 {\kb}		{{k_\mathrm{B}}}
\newcommand	 {\Tlow}	{{T_\mathrm{low}}}
\newcommand 	 {\Pnat} 	{{P_\mathrm{nat}}}
\newcommand {\QIA}	{{Q_\mathrm{IA}}}
\newcommand {\QIB}	{{Q_\mathrm{IB}}}
\newcommand {\QIIA}	{{Q_\mathrm{IIA}}}
\newcommand {\QIIB}	{{Q_\mathrm{IIB}}}
\newcommand {\Pcut}     	{{P_\mathrm{cut}}}
\newcommand {\TlowI}     {{T^\mathrm{I}_\mathrm{low}}}
\newcommand {\TlowII}    {{T^\mathrm{II}_\mathrm{low}}}
\newcommand {\Ptot}	{{P_\mathrm{tot}}}
\newcommand {\PIA}    	{{P_\mathrm{IA}}}
\newcommand {\PIB}    	{{P_\mathrm{IB}}}
\newcommand {\PIIA}    	{{P_\mathrm{IIA}}}
\newcommand {\PIIB}    	{{P_\mathrm{IIB}}}
\newcommand {\Eave}	{{E_\mathrm{ave}}}
\newcommand {\sigE}	{{\sigma_{\left < E \right >}}}
\newcommand {\SR}		{${\mathrm{S16}_{144}}$}
\newcommand {\SI}		{${\mathrm{S16}_{1024}}$}	
\newcommand {\SII}		{${\mathrm{S35}_{1024}}$}

@@

\begin{abstract}
We consider a generalized-ensemble algorithm for coarse-grained simulations of biomolecules which allows the thermodynamic behavior of two or more sequences to be determined in a single ``multisequence" run. By carrying out a random walk in sequence space, the method also enhances conformational sampling. Escape from local energy minima is accelerated by visiting sequences for which the minima are more shallow or absent. We apply the method alongside an intermediate-resolution coarse-grained model for protein folding with 7 atoms per amino acid and 3 amino acid types. The potential for large-scale coverage of sequence space is explored by applying the method to two sets with $>$1,000 sequences each. The resulting thermodynamic data is used to explore the sequence-structure and sequence-stability relationships between pairs of protein folds with different secondary structures.
\end{abstract}

\pacs{87.14.E; 87.15.A; 05.10.Ln}

\keywords{Monte Carlo, generalized ensembles, simulated tempering, protein folding, protein evolution}

@@

\section{Introduction}
\noindent
Recent years have seen important advances in biomolecular simulation methods, including improvements to standard molecular dynamics force fields,~\cite{Piana2014} the advent of several alternative atomistic simulation approaches,~\cite{Ding2008,Irback2006,Verma2009,Yang2007} and new techniques for  conformational sampling.~\cite{Bernardi2015} These advances have in part triggered major efforts~\cite{McGuffee2010,Perilla2016,Lindorff-Larsen2011,Yu2016} to characterize the molecular dynamics of bio-systems of different sizes, e.g., a small native protein on the millisecond scale~\cite{Lindorff-Larsen2011} or a comprehensive model cytoplasm on the nanosecond scale.~\cite{Yu2016} While encouraging and insightful, these large-scale simulations have also highlighted the fact that severe tradeoffs in size and time scales will likely persist for the foreseeable future for detailed simulations.

One way to expand the range of biomolecular simulations is to turn to coarse-grained (CG) models, where the basic aim is to simplify the physical description of interactions while retaining the essential physics of the system under study.~\cite{Riniker2012} Ingolfsson \textit{et al.} list 4 main factors that make CG models computationally fast: reduction in the number of degrees of freedom, faster simulation dynamics, emphasis on short-range interactions and the ability of using larger integration time steps.~\cite{Ingolfsson2014} To this list can be added that a CG representation either of the interaction potential or the molecular geometry often opens up for alternative sampling schemes beyond traditional molecular dynamics approaches which can further speed up conformational sampling, e.g., activation-relaxation kinetics,~\cite{Beland2011} discrete molecular dynamics~\cite{Proctor2011} and various Monte Carlo (MC)-based techniques such as cluster moves.~\cite{Vitalis2009}

The challenges of achieving representative conformational sampling of individual biomolecular systems notwithstanding, a range of systems naturally call for the investigation and comparison of many molecular variants, e.g., determining the molecular mechanisms of specificity in protein-protein~\cite{Zarrinpar2003,Hakes2007} or protein-nucleotide interactions,~\cite{Rohs2010} or the role of mutations in disease processes such as protein aggregation.~\cite{Ross2004} Another example is protein folding where unique insight has been achieved by comparing the folding within and between protein families.~\cite{Tzul2017,Wensley2010} With the increasing availability of sequence information,~\cite{Vukmirovic2000} it is of interest to explore ways to efficiently sample multiple sequences in biomolecular simulations.

To this end, we evaluate in this work an MC-based algorithm that can in principle calculate the thermodynamics of multiple sequences in a single run and apply it to a coarse-grained model for protein folding. This ``multisequence" (MS) method was originally developed in the context of homo- and heteropolymer simulations~\cite{Irback1995} and was later adapted for the characterization of peptide-protein binding specificity,~\cite{Bhattacherjee2013,Wallin2017}, but it has to our knowledge not been previously tested in protein folding simulations. The MS algorithm relies on sampling at fixed temperature a generalized ensemble in which the sequence is a dynamic parameter. Hence, there are two main types of updates: conformational updates $\rbar\rightarrow\rbar'$ and sequence updates $\sbar\rightarrow\sbar'$. As illustrated in Fig.~1, this approach is easily applied when $\rbar$ and $\sbar$ are ``perpendicular" coordinates such that the potential energy of the model can be written in terms of two independent variables, $E(\sbar,\rbar)$. While this may be hard to achieve for detailed protein models, it is likely possible in many CG models.

As an initial test case, we apply the MS method to a set of 144 model sequences with 16 amino acids. This set, {\SR}, was constructed to sparsely cover the sequence space between two ideally designed sequences, A1 and N1, which fold into an $\alpha$-helix and $\beta$-hairpin, respectively, and allowed us to demonstrate the existence of A1-N1 mutational pathways which abruptly switch between the two structures and do not proceed through unstable sequences.~\cite{Holzgrafe2014} Here we use this set to validate the MS method and compare its efficiency to a standard generalized-ensemble method, namely simulated tempering (ST).~\cite{Marinari1992,Lyubartsev1992} We thereafter apply the MS method to a greatly enlarged set with 1,024 sequences, also spanning the A1-N1 space, as well as another set spanning two 35-amino acid sequences, A2 and TN, that fold into two-helical bundle and mixed $\alpha$-$\beta$ structures, respectively. Besides hinting at the feasibility of applying the MS method on large sets of sequences, the results allow us to carry out a more systematic analysis of the biophysical properties of sequences along mutational pathways connecting these two pairs of basic protein folds than has been possible with standard methods.~\cite{Holzgrafe2014,Holzgrafe2015}

@@

\subsection{Model sequences}
\noindent
Six of the model sequences studied in this work, A1, N1, R1, R2, A2, and TN, are given in Table~I. In addition, we study two sequence sets $\mathrm{S16}_{1024}$  and $\mathrm{S35}_{1024}$ with 1,024 sequences each derived from the A1-N1 and A2-TN pairs, respectively, through mutational combinations, as well as the set $\mathrm{S16}_{144}$ taken from  Ref.~\protect\citenum{Holzgrafe2014}.

@@

\subsection{Monte Carlo simulation parameters}
\noindent
Both ST and MS simulations are carried out with two types of conformational updates $\rbar\rightarrow\rbar'$: (1) a global pivot move (20\%) which randomly picks a $\phi_\mathrm{i}$ angle or $\psi_\mathrm{i}$ angle and assign a new value between $-\pi$ and $\pi$; and (2) a semi-local move (80\%) which turns the $\phi_\mathrm{i}$ and $\psi_\mathrm{i}$-angles of 4 consecutive amino acids in a coordinated manner.~\cite{Favrin2001} In MS simulations, sequence updates $\sbar\rightarrow\sbar'$ are carried out by randomly picking a new sequence $\sbar'\ne\sbar$ and applying the accept-reject criterion in Eq.~\ref{accrej}. A sequence move is attempted every $1,000$ MC steps while temperature updates $m\rightarrow m'$ are attempted every 100 steps. The simulations carried out in this work are summarized in Table~II.

\begin{table}
\caption{\label{tab2} List of simulations carried out in this work. }
\begin{ruledtabular}
\begin{tabular}{lcccr}
Runs & Algorithm & $\kb T$  & MC steps/run\footnote{Excludes a thermalization step with $10^6$ MC steps/run.} &  Sequences\\
\hline
32 & ST & 0.43--0.65 & $1\times10^7$ &A1\\ 
32 & ST & 0.43--0.65 & $1\times10^7$ &N1\\ 
32 & ST & 0.43--0.65 & $1\times10^7$ &R1\\ 
32 & ST & 0.43--0.65 & $1\times10^7$ &R2\\ 
32$\times$8\footnote{32 runs per temperature at 8 different temperatures.} & MS &0.43--0.65& $18\times 10^7$ & $\mathrm{S16}_{144}$\\
16 & MS & 0.43  & $5\times 10^9$ &  $\mathrm{S16}_{1024}$ \\
16 & MS & 0.46 & $4\times 10^9$ &  $\mathrm{S35}_{1024}$ \\
\end{tabular}
\end{ruledtabular}
\end{table}

\subsection{Observables}
\noindent
Fold stabilities are calculated as in Ref.~\citenum{Holzgrafe2015} and described briefly below. First we define two structural similarity measures $\QIA$ and $\QIB$ for folds IA and IB, respectively, indicating the fraction of the fold-specific backbone-backbone hydrogen bonds that have been formed. The fold IA-hydrogen bonds are (2,6), (3,7), (4,8), (5,9), (6,10), (7,11), (8,12), (9,13), (10,14), (11,15) and the fold IB-bonds are (3,14), (5,12), (7,10), (10,7), (12,5), (14,3), where (i,j) indicates a hydrogen bond between the CO group of amino acid i and the NH group of amino acid j. The  stabilities of folds IA and IB are then defined as the probabilities $\PIA = P(\QIA\ge0.8)$ and $\PIB = P(\QIB\ge0.80)$, respectively, i.e., the probability that at least 80\% of the fold's hydrogen bonds are formed. $\PIA$ and $\PIB$ thus depend on both sequence $\sbar$ and temperature $T$. For example, $\PIA=0.875\pm 0.003$ for A1 and $\PIB=0.785\pm0.008$ for N1 at $\kb T = 0.43$. Structural similarity measures for 35-amino acid folds IIA and IIB are defined as $\QIIA = ( Q_\mathrm{IA}^\mathrm{1-16} + Q_\mathrm{IA}^\mathrm{20-35} + Q_\mathrm{tert} ) / 3$ and $\QIIB = ( Q_\mathrm{IA}^\mathrm{1-16} + Q_\mathrm{IB}^\mathrm{20-35} + Q_\mathrm{tert} ) / 3$, respectively, where superscripts on $\QIA$ and $\QIB$ indicate over which amino acid positions those measures are applied to within the longer 35 amino acid sequences and $Q_\mathrm{tert}$ is a measure that counts the number of $\mathrm{C}_\beta$-$\mathrm{C}_\beta$ contacts between the two secondary structure elements of these folds.~\cite{Holzgrafe2015} In analogy with $\PIA$ and $\PIB$, we define the stabilities of folds IIA and IIB as $\PIIA = P(\QIIA\ge0.8)$ and $\PIIB = P(\QIIB\ge0.80)$, respectively. The root-mean-square-deviation, RMSD, is calculated over all $\mathrm{C}_\alpha$ atoms.

\begin{figure}
\includegraphics[width=10.0cm]{MCTrajFolds}
\caption{(A) Example of an MS simulation applied to the sequence set $\mathrm{S16}_{144}$. The plot shows the MC evolution of the sequence $\sbar$ (numbered 1--144), the total potential energy $E$ and the root-mean-square deviation (RMSD) calculated against fold IA (light blue) and fold IB (dark red) structures shown in (B). The simulation is carried out at $\kb T = 0.43$. (B) Representative structures of fold IA and fold IB, selected to be the minimum-energy conformations found for the sequences A1 and N1, respectively. (C) Representative fold IIA and fold IIB structures selected to be the minimum-energy conformations of A2 and TN, respectively. }
\end{figure}

@@

\subsection{Computational efficiency of the multisequence algorithm}
\noindent
We start by testing the MS algorithm on the set {\SR} consisting of the 144 3-letter model sequences with 16 amino acids studied in Ref.~\citenum{Holzgrafe2014}. Two of the sequences in {\SR} are A1 and N1 (see Table~I) which were designed to fold into stable $\alpha$-helix and $\beta$-hairpin structures, respectively, as shown in Fig.~2B. A1 and N1 differ at 10 positions such that 10 consecutive point mutations can transform A1 into N1, and vice versa. The binary sequence space between A1 and N1 in which any combination of these mutations have been carried out, therefore contains $2^{10}=1,024$ sequences. The 144 sequences were selected from this binary space with the constraint that the total number of h-amino acids are not too high and not too unevenly distributed.~\cite{Holzgrafe2014}

\begin{figure}
\includegraphics[width=7.8cm]{Pacc}
\caption{Acceptance rates for $\sbar\rightarrow\sbar'$ updates in MS simulations of the {\SR} sequence set as a function of (A) the number of changed amino acid positions $\Delta h$ and (B) temperature $T$. Acceptance rates for 3 different $T$'s are shown in (A).}
\end{figure}

We apply the multisequence algorithm on the set {\SR} across a range of temperatures (see Table~II). Figure~2 illustrates a typical MS simulation trajectory carried out at the lowest temperature which is well below the folding temperature of both A1 and N1~\cite{Holzgrafe2014,Holzgrafe2015}. From the MC evolution of the total energy $E$, sequence index $\sbar$, and RMSD values from the representative structures in Fig.~2B, it is evident that visits to various sequences drive transitions into a range of structural states. In particular, there are frequent visits to both $\alpha$-helix and $\beta$-hairpin structures. Transitions between them are accompanied by a shift in which sequences are preferably visited. For example, visits to high $\sbar$-indices, including N1 with index 144, tend to coincide with formation of $\beta$-hairpin structures.

One might have suspected that the MS algorithm would be hampered by poor acceptance rates for sequence updates. However, this is not the case. We carry out updates $\sbar\rightarrow\sbar'$ by picking a new random sequence $\sbar'\ne\sbar$ from the set of allowed sequences. The (average) acceptance rate  depends on both $T$ and the step in sequences space $\Delta h$, i.e., the number of amino acid positions changed, as shown in Fig.~3. At the lowest $T$ and high $\Delta h$, acceptance rates are only around 0.1-0.2 but are substantially higher at other $T$ and $\Delta h$ values. The overall acceptance rates are in fact above the oft-quoted rule-of-thumb value $0.25$~\cite{Gilks1996} for all temperatures $T$ (see Fig~4B). Increased acceptance rates can easily be achieved by restricting proposed updates such that $\Delta h\le\Delta h_\mathrm{max}$, where $\Delta h_\mathrm{max}$ is a maximum step size, which might be necessary for longer protein chains.

\begin{figure}
\includegraphics[width=8.0cm]{Stderr}
\caption{Comparing sampling efficiency of the MS and ST algorithms. Statistical errors $\sigma_{\left < E\right >}$ of the average total energy $\left < E\right >$ from the two methods for the 16-amino acid sequences (A) A1, (B) N1, (C) R1 and (D) R2. Simulation lengths in the two methods are adjusted such that the number of conformations sampled per sequence and temperature is roughly the same (see text). }
\end{figure}

We now compare the results from our MS calculations with simulated tempering (ST) simulations carried out on 4 of the 144 sequences, namely A1 and N1 and two sequences, R1 and R2, chosen randomly at distances $h=4$ and $h=6$ from A1, respectively (see Table~I). While ST provides the thermodynamics of a given sequence across a range of $T$ in a single run, a MS simulation provides the thermodynamics of all 144 sequences for one $T$. We adjust the simulation lengths for ST and MS runs such that roughly the same number of sampled conformations are obtained for each $\sbar$ and $T$ combination, thus ensuring that similar computational resources are used for the two algorithms (see Table~II). The MS algorithm is validated by comparing the average total energy, $\left < E \right >$, calculated for different $T$'s and $\sbar$'s, between the two sets of results (see Supplementary Information).

As a way to assess sampling efficiency, we compare in Figure~4 the statistical error, $\sigE$, of the average energy $\left <E\right >$ for the 4 different sequences obtained using ST and MS, respectively. Since roughly the same number of sampled conformations were obtained for each s and $T$ combination, we can compare the statistical errors directly. Both MS and ST algorithms give the same small error at the highest $T$ studied which implies consistent result between them. It should be noted that the highest $T$ studied is well above the folding temperature for all the 4 sequences. Therefore, this consistent result is not particularly surprising since at high-T the free-energy landscape is relatively smooth. Hence, conformational space requires little or no difficulty to sample.

@@

Insight from the statistical measurements suggests that the MS algorithm has a similar computational efficiency at high $T$ and even better at lower $T$'s. The MS algorithm is able to achieve  better efficiency by sampling a wider region of both conformational and sequence space. Sampling the sequence space in addition to conformational space ensures a decorrelated ensemble of states is generated.

\begin{figure*}
\rotatebox{0}{\includegraphics[width=16.6cm]{network}}
\caption{Networks of sequences connecting folds IA and IB (top) and folds IIA and IIB (bottom). Each node represents a stable sequence ($\Ptot\ge\Pcut$ where $\Pcut=0.50$) that folds into either IA or IIA (light blue), IB or IIB (dark red), or is classified as bistable ($B>0.5$, black). A line between two nodes indicates that the sequences differ at only one position. Graph created using the tool Graphviz~\protect\cite{Graphviz2000} obtained from  www.graphviz.org.}
\end{figure*}
